---
title: "Example code for GAMM talk at PaPE 2017 stats workshop"
author: "Márton Sóskuthy"
date: "14/06/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

This file is a supplement to a talk that I gave at a satellite workshop on statistical modelling in phonetics at PaPE 2017 in Köln. The code in this file illustrates the structure of the statistical models that I tested for type I error and power. Only a few example models are shown in this file, but the rest of the models covered in the talk are easy to derive from these.

The data set is generated randomly by sourcing an external script, which was used as part of the simulations. The code for the type I / II error simulations as well as the raw data are also available in this GitHub repo, but they are documented separately. Some supporting functions for analysing the results are also provided, though these are not documented in detail.

## Setting things up

Loading the libraries & creating the data set using a separate script.

```{r}
library(mgcv)
library(itsadug)
library(ggplot2)

source("formants/formants_type_2_init.R")
```

For illustration, here's what the data set looks like:

```{r}
head(dat)
ggplot(dat, aes(x=measurement.no, y=f2, col=group, group=traj)) + geom_line()
```

## Varying model structure, significance testing based on model summary

Here's a simple GAM without any random structures using a difference smooth to represent the difference between the two groups of trajectories. Note that the data set is already set up so that the grouping variable (group.ordered) is an ordered factor with contrast coding.

```{r}
mod.noranef <- bam(f2 ~ group.ordered +                        # parametric term
                        s(measurement.no) +                    # reference smooth, bs="tp", k=10
                        s(measurement.no, by=group.ordered),   # difference smooth, bs="tp", k=10
                   data=dat)
summary(mod.noranef)
```

Here are some further elaborations on this model using different random structures.

```{r}
# random intercepts by trajectory
mod.rintcpt <- bam(f2 ~ group.ordered +
                        s(measurement.no) +
                        s(measurement.no, by=group.ordered) +
                        s(traj, bs="re"),
                   data=dat)
summary(mod.rintcpt)

# random intercepts + slopes by trajectory
mod.rslope <- bam(f2 ~ group.ordered +
                       s(measurement.no) +
                       s(measurement.no, by=group.ordered) +
                       s(traj, bs="re") + s(traj, measurement.no, bs="re"),
                  data=dat)
summary(mod.rslope)

# random smooths by trajectory (smooth type set by "xt"; k=10)
# warning about repeated 1-d smooths can be ignored safely
mod.rsmooth <- bam(f2 ~ group.ordered +
                        s(measurement.no) +
                        s(measurement.no, by=group.ordered) +
                        s(measurement.no, traj, bs="fs", m=1, xt="tp", k=10),
                   data=dat)
summary(mod.rsmooth)
```

And a version with an autoregressive error model, where the degree of autocorrelation is estimated from a simple version of the model (in this case, mod.noranef). Note that the data set contains a column that indicates the start of each trajectory using the value TRUE (also, points from within the same trajectory are adjacent and ordered according to measurement.no).

```{r}
# step 1: extracting AR coefficient from mod.noranef
mod.noranef.rho <- start_value_rho(mod.noranef)

# step 2:
mod.AR <- bam(f2 ~ group.ordered +
                   s(measurement.no) +
                   s(measurement.no, by=group.ordered),
              data=dat,
              start.value=dat$start,
              rho=mod.noranef.rho)
summary(mod.AR)
```

## Other methods of significance testing

Significance testing using model comparison where both the parametric term and the difference smooth are excluded at the same time. Note the use of maximum likelihood (ML) for parameter estimation.

```{r}
# full model
mod.AR.ML <- bam(f2 ~ group.ordered +
                      s(measurement.no) +
                      s(measurement.no, by=group.ordered),
                 data=dat,
                 start.value=dat$start,
                 rho=mod.noranef.rho,
                 method="ML")
summary(mod.AR.ML)

# nested model
mod.AR.ML.nested <- bam(f2 ~ s(measurement.no),
                        data=dat,
                        start.value=dat$start,
                        rho=mod.noranef.rho,
                        method="ML")

# model comparison

compareML(mod.AR.ML, mod.AR.ML.nested)
```

Visual comparison using confidence intervals around model predictions.

```{r}
plot_smooth(mod.rsmooth, view="measurement.no", plot_all="group.ordered",
            rm.ranef=T, rug=F)
```

Visual comparison using difference smooth.

```{r}
plot_diff(mod.rsmooth, view="measurement.no", comp=list(group.ordered=c("B","A")),
            rm.ranef=T)
```