---
title: "Models for set 1, pitch results (GAMM modelling strategies)"
author: "Márton Sóskuthy"
date: "25/05/2018"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document is a supplement to "Evaluating generalised additive mixed modelling strategies for dynamic speech analysis," relating specifically to the contents of Table 2 in Section 3.1. It presents code that illustrates (i) how the simulated data were generated and (ii) the models whose performance is summarised Table 1.

## Preliminaries

The code below loads the relevant libraries.

```{r, message=F}
library(ggplot2)
library(mgcv)
library(itsadug)
```

## Data generation

The code in this section can be used to create data for either type I or type II simulations. Set the value of *type* to 1 for type I simulations and to 2 for type II simulations. To replicate the results in the table in appendix C, which summarises the same types of trajectories, but with a small amount of added noise, set the value of *noise* to TRUE.

```{r}
type = 1
noise = FALSE
```

The data for this set of simulations consist of simulated pitch trajectories loosely modelled after triconstituent compounds in English. 50 trajectories are generated. For type I simulations, these are randomly assigned to two groups (A and B). For type II simulations, all group B trajectories are slightly modified (cf. Section 2.1 in the paper and also the Appendix).

The following code sets the parameters that determine the main characteristics of the data set.

```{r}
# setting time dimension (these are the points along which the trajectories are generated)
xs_dense = seq(0,1,0.025)

# indices for extracting every second measurement; this is used for downsampling
xs_thin_ind = c(rep(c(T,F), (length(xs_dense)-1)/2), T)

# expected values & sd for starting and end points
start_mean = 170
start_sd = 10
slope_mean = -30
slope_sd = 8

# boundaries between N1, N2 and N3

boundary.1_min = (1/3) - 0.1
boundary.1_max = (1/3) + 0.1

boundary.2_min = (2/3) - 0.1
boundary.2_max = (2/3) + 0.1

# pitch accent 1 - vertical target
H.star.1_vertical_mean = 15 # roughly the equivalent of ~ 1.5-2 semitones
H.star.1_vertical_sd = 6

# pitch accent 1 - horizontal target
H.star.1_horizontal_min = 0 # as a proportion of the duration of W1
H.star.1_horizontal_max = 0.25

# pitch accent 1 - width
H.star.1_bw = 0.12 # as a proportion of *overall* duration

# pitch accent 2 - vertical target
H.star.2_vertical_mean = 4.5
H.star.2_vertical_sd = 6

# pitch accent 2 - horizontal target
H.star.2_horizontal_min = 0 # as a proportion of the duration of W2
H.star.2_horizontal_max = 0.25

# pitch accent 2 - width
H.star.2_bw = 0.12 # as a proportion of *overall* duration
  
# boundary tone - vertical target
H.minus_vertical_mean = 8 # about half of H.star
H.minus_vertical_sd = 3

# boundary tone - width
H.minus_bw = 0.08

# final boundary tone - vertical target & width
L.percent_vertical = -20
L.percent_bw = 0.12

# noise
noise_sd = 1

# number of trajectories to generate
n_trajectories <- 50

```

The code below assembles the dense version of the data set.

```{r}
# creating matrix that will store the trajectories
ys_m <- matrix(0, nrow=length(xs_dense), ncol=n_trajectories)

# assembling individual trajectories in pairs (one for group A and one for group B)
for (i in 1:n_trajectories) {
  # sampling start values, slopes, boundary positions and pitch accents / boundary tones 
  # for trajectories A and B
  start <- rnorm(1, start_mean, start_sd)
  slope <- rnorm(1, slope_mean, slope_sd)
  boundary.1 <- runif(1, boundary.1_min, boundary.1_max)
  boundary.2 <- runif(1, boundary.2_min, boundary.2_max)
  H.star.1_horizontal <- runif(1, H.star.1_horizontal_min, H.star.1_horizontal_max)
  H.star.1_vertical <- rnorm(1, H.star.1_vertical_mean, H.star.1_vertical_sd)
  H.star.2_horizontal <- runif(1, H.star.2_horizontal_min, H.star.2_horizontal_max)
  H.star.2_vertical <- rnorm(1, H.star.2_vertical_mean, H.star.2_vertical_sd)
  H.minus_vertical <- rnorm(1, H.minus_vertical_mean, H.minus_vertical_sd)
  
  # assembling the trajectories
  ys_m[,i] <- start + xs_dense*slope +  # declination
    exp(-((xs_dense - (boundary.1*H.star.1_horizontal))**2)/(2*H.star.1_bw**2)) * H.star.1_vertical + # 1st pitch accent
    exp(-((xs_dense - boundary.2)**2)/(2*H.minus_bw**2)) * H.minus_vertical + # boundary tone
    exp(-((xs_dense - 1)**2)/(2*L.percent_bw**2)) * L.percent_vertical # final boundary tone
  
  # adding additional pitch accent for group B if this is a type II simulation
  if (i > (n_trajectories/2) & type==2) {
    ys_m[,i] <- ys_m[,i] + 
      exp(-((xs_dense - (boundary.1+(boundary.2-boundary.1)*H.star.2_horizontal))**2)/(2*H.star.2_bw**2)) * H.star.2_vertical # 2nd pitch accent
  }
  
  # adding noise
  if (noise) {
    ys_m[,i] <- ys_m[,i] + rnorm(length(xs_dense), 0, noise_sd)
  }
}

# assembling data set
dat_dense <- data.frame(traj=paste("traj_", rep(1:n_trajectories, each=length(xs_dense)), sep=""), # traj ids
                  group=rep(c("A","B"), each=length(xs_dense)*(n_trajectories / 2)),               # group ids
                  measurement.no=xs_dense,                          # measurement number (i.e. time dimension)
                  pitch=c(ys_m),                                                             # pitch  values
                  stringsAsFactors = F
)

# setting up different types of grouping factors
dat_dense$group.factor <- as.factor(dat_dense$group)
dat_dense$group.ordered <- as.ordered(dat_dense$group) 
contrasts(dat_dense$group.ordered) <- "contr.treatment"
dat_dense$group.bin <- as.numeric(dat_dense$group.factor) - 1

# trajectory ids must be factors  
dat_dense$traj <- as.factor(dat_dense$traj)

# add dat$start for AR.start (for autoregressive error models)
dat_dense$start <- dat_dense$measurement.no == 0
```

Downsampling to thin version of the data set.

```{r}
dat_thin <- dat_dense[rep(xs_thin_ind, n_trajectories),]
```

Here is what the data set looks like.

```{r}
ggplot(dat_dense, aes(x=measurement.no, y=pitch, group=traj, col=group)) +
  geom_line() +
  facet_grid(~group)
```

## Models

All the models (and sets of models) from Table 1 are shown below in the same order as in the table.

### 1. No components

```{r}
# dense
nocomp_dense <- bam(pitch ~ group.ordered + 
                      s(measurement.no, bs = "tp", k = 15) + 
                      s(measurement.no, by = group.ordered, bs = "tp", k = 15),
                    data = dat_dense, method = "fREML", discrete = T, nthreads = 1)
summary(nocomp_dense)

# thin
nocomp_thin <- bam(pitch ~ group.ordered + 
                     s(measurement.no, bs = "tp", k = 15) + 
                     s(measurement.no, by = group.ordered, bs = "tp", k = 15), 
                   data = dat_thin, method = "fREML", discrete = T, nthreads = 1)
summary(nocomp_thin)
```

### 2. Rand intcpt (= Random Intercept)

```{r}
# dense
rand_intcpt_dense <- bam(pitch ~ group.ordered + 
                           s(measurement.no, bs = "tp", k = 15) + 
                           s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
                           s(traj, bs = "re"), 
                         data = dat_dense, method = "fREML", discrete = T, nthreads = 1)
summary(rand_intcpt_dense)

# thin
rand_intcpt_thin <- bam(pitch ~ group.ordered + 
                           s(measurement.no, bs = "tp", k = 15) + 
                           s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
                           s(traj, bs = "re"), 
                         data = dat_thin, method = "fREML", discrete = T, nthreads = 1)
summary(rand_intcpt_thin)
```

### 3. Rand intcpt + slope (= random intercept + slope)

```{r}
# dense
rand_intcpt_slope_dense <- 
  bam(pitch ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 15) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
        s(traj, bs = "re") + 
        s(traj, measurement.no, bs = "re"), 
      data = dat_dense, method = "fREML", discrete = T, nthreads = 1) 
summary(rand_intcpt_slope_dense)

# thin
rand_intcpt_slope_thin <- 
  bam(pitch ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 15) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
        s(traj, bs = "re") + 
        s(traj, measurement.no, bs = "re"), 
      data = dat_thin, method = "fREML", discrete = T, nthreads = 1) 
summary(rand_intcpt_slope_thin)
```

### 4. Rand smooth, 5 bs

```{r}
# dense
rand_smooth_3_dense <- 
  bam(pitch ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 15) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
        s(measurement.no, traj, bs = "fs", m = 1, xt = "tp", k = 5), 
      data = dat_dense, method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_3_dense)

# thin
rand_smooth_3_thin <- 
  bam(pitch ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 15) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
        s(measurement.no, traj, bs = "fs", m = 1, xt = "tp", k = 5), 
      data = dat_thin, method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_3_thin)
```

### 5. Rand smooth, 8 bs

```{r}
# dense
rand_smooth_5_dense <- 
  bam(pitch ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 15) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
        s(measurement.no, traj, bs = "fs", m = 1, xt = "tp", k = 8), 
      data = dat_dense, method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_5_dense)

# thin
rand_smooth_5_thin <- 
  bam(pitch ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 15) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
        s(measurement.no, traj, bs = "fs", m = 1, xt = "tp", k = 8), 
      data = dat_thin, method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_5_thin)
```

### 6. Rand smooth, 12 bs

```{r}
# dense
rand_smooth_10_dense <- 
  bam(pitch ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 15) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
        s(measurement.no, traj, bs = "fs", m = 1, xt = "tp", k = 12), 
      data = dat_dense, method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_10_dense)

# thin
rand_smooth_10_thin <- 
  bam(pitch ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 15) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
        s(measurement.no, traj, bs = "fs", m = 1, xt = "tp", k = 12), 
      data = dat_thin, method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_10_thin)
```

### 7. Rand smooth, gam.check

The model fitted in this section uses random smooths where the number of basis functions (k) is determined using the gam.check() function: after fitting an initial model with a relatively low value of k, gam.check() is used to see whether more wiggliness is necessary (essentially, whether the smooths use up all the degrees of freedom afforded to them). If gam.check() suggests that more wiggliness is necessary, this procedure is repeated again using a model with a higher value of k.

Below is a convenience function for extracting the relevant p-value from the output of gam.check.

```{r}
gam.check.p.value <- function (mod, which.line) { # which.line is a regexp
  str.out <- capture.output(gam.check(mod))
  relevant.line <- str.out[grep(which.line, str.out)]
  p.value <- as.numeric(gsub(".*?([0-9.e-]*)[ .*]*$", "\\1", relevant.line, perl=T))
  return(p.value)
}
```

Fitting the models. Dense first.

```{r}
# what k's should be tried?
k_min = 4
k_max = 12
k_step = 4
```

```{r, fig.keep='none'}
# dense
for (k in seq(k_min,k_max,k_step)) {
  cat("fitting model with  k =", k, "\n")
  
  # fitting model
  
  rand_smooth_gam.check_dense <- 
    bam(pitch ~ group.ordered + 
          s(measurement.no, bs = "tp", k = 15) + 
          s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
          s(measurement.no, traj, bs = "fs", m = 1, xt = "tp", k = k), 
        data = dat_dense, method = "fREML", discrete = T, nthreads = 1)
  
  # check whether more complexity is needed using gam.check
  
  if (gam.check.p.value(rand_smooth_gam.check_dense, "traj") >= 0.05 | k == k_max) {
    print(summary(rand_smooth_gam.check_dense))
    break
  }
}
``` 

And now the thin data set.

```{r, fig.keep='none'}
# thin
for (k in seq(k_min,k_max,k_step)) {
  cat("fitting model with  k =", k, "\n")
  
  # fitting model
  
  rand_smooth_gam.check_thin <- 
    bam(pitch ~ group.ordered + 
          s(measurement.no, bs = "tp", k = 15) + 
          s(measurement.no, by = group.ordered, bs = "tp", k = 15) + 
          s(measurement.no, traj, bs = "fs", m = 1, xt = "tp", k = k), 
        data = dat_thin, method = "fREML", discrete = T, nthreads = 1)
  
  # check whether more complexity is needed using gam.check
  
  if (gam.check.p.value(rand_smooth_gam.check_thin, "traj") >= 0.05 | k == k_max) {
    print(summary(rand_smooth_gam.check_thin))
    break
  }
}
```

### 8. AR1

First fitting models without AR component in order to estimate rho. This is equivalent to the nocomp model above.

```{r}
# dense
nocomp_dense <- bam(pitch ~ group.ordered + 
                      s(measurement.no, bs = "tp", k = 15) + 
                      s(measurement.no, by = group.ordered, bs = "tp", k = 15), 
                    data = dat_dense, method = "fREML", discrete = T, nthreads = 1)
# thin
nocomp_thin <- bam(pitch ~ group.ordered + 
                      s(measurement.no, bs = "tp", k = 15) + 
                      s(measurement.no, by = group.ordered, bs = "tp", k = 15), 
                    data = dat_thin, method = "fREML", discrete = T, nthreads = 1)
```

Extracting rho.

```{r}
rho_dense <- start_value_rho(nocomp_dense)
cat("rho =", rho_dense, "for the dense data set\n")
rho_thin <- start_value_rho(nocomp_thin)
cat("rho =", rho_thin, "for the thin data set\n")
```

Fitting models with AR1

```{r}
# dense
AR1_dense <- bam(pitch ~ group.ordered + 
                   s(measurement.no, bs = "tp", k = 15) + 
                   s(measurement.no, by = group.ordered, bs = "tp", k = 15), 
                 data = dat_dense, 
                 AR.start = dat_dense$start, rho = rho_dense, 
                 method = "fREML", discrete = T, nthreads = 1)
summary(AR1_dense)

# thin
AR1_thin <- bam(pitch ~ group.ordered + 
                   s(measurement.no, bs = "tp", k = 15) + 
                   s(measurement.no, by = group.ordered, bs = "tp", k = 15), 
                 data = dat_thin, 
                 AR.start = dat_thin$start, rho = rho_thin, 
                 method = "fREML", discrete = T, nthreads = 1)
summary(AR1_thin)
```