---
title: "GAMM Modelling Strategies paper (JPhon): Analysis"
output: html_notebook
---

```{r "setup", include=FALSE}
require("knitr")
# setting workind dir
opts_knit$set(root.dir = "~/documents/research/projects/dynamic-gam/gamm_modelling_strategies")
```

Importing relevant libraries & main data set.

```{r}
library(mgcv)
library(itsadug)
library(data.table)
library(stringr)
library(xtable)

dat <- readRDS("gamm_sim_master.rds")
```

## AR models & random effects for capturing autocorrelation

First, let's have a look at how well different types of random structures and AR models do at capturing autocorrelation within (relatively) smooth trajectories. We'll only look at a subset of the data at this point: `sim.type == "simple"`. We'll go straight for a well-formatted table.

Formants first.

```{r}
dat.simple.formants <- dat[dat$sim.type=="simple" & dat$data.type=="formants",]
# aggregating data
dat.simple.formants.aggr <- 
  dat.simple.formants[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                         par.type.1=mean(par.p[error.type=="type_1"] < 0.05),
                         smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                         par.power=mean(par.p[error.type=="type_2"] < 0.05),
                         time.med=median(time),
                         memory.med=median(memory)),
                      by=.(data.density,ranef,AR)]
kable(dat.simple.formants.aggr)

# creating output table for formant data
# record timing / memory for dense data, set aside
sf.time.mem <- dat.simple.formants.aggr[data.density=="dense",c("ranef","AR","time.med","memory.med")]
# go to wide format
sf.out <- dcast.data.table(dat.simple.formants.aggr, ranef + AR ~ data.density, value.var=c("par.type.1","smooth.type.1","par.power","smooth.power"))
# reorder rows
sf.out <- sf.out[c(2:7,1),]
# merge back with timing / memory stuff
sf.out <- sf.out[sf.time.mem, on=.(ranef,AR)]
# new column that codes ranef and AR together
sf.out[,model.component:=ifelse(AR=="AR_est","AR1",ranef)]
sf.out$model.component <- recode(sf.out$model.component,
                                 noranef="no components",
                                 rintcpt="rand intcpt",
                                 rslope="rand intcpt + slope",
                                 rsmooth_tp_3="rand smooth, 3 bs",
                                 rsmooth_tp_5="rand smooth, 5 bs",
                                 rsmooth_tp_10="rand smooth, 10 bs")
sf.out <- sf.out[,-(1:2)]
setcolorder(sf.out, c("model.component",
                      "par.type.1_thin","par.power_thin",
                      "smooth.type.1_thin","smooth.power_thin",
                      "par.type.1_dense","par.power_dense",
                      "smooth.type.1_dense","smooth.power_dense",
                      "memory.med","time.med"))
colnames(sf.out) <- c("model component",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "memory","time")
sf.out <- as.data.frame(sf.out)
for (i in seq(2,8,2)) {
  sf.out[,i] <- as.character(round(sf.out[,i],3))
  sf.out[,i+1] <- as.character(round(sf.out[,i+1],3))
  for (j in 1:nrow(sf.out)) {
    if (as.numeric(sf.out[j,i]) > 0.15) {
      sf.out[j,i] <- paste0("\\color{lightgray} ",sf.out[j,i])
      sf.out[j,i+1] <- paste0("\\color{lightgray} ",sf.out[j,i+1])
    }
  }
}
sf.str.out <- print(xtable(sf.out, digits=c(0,0,rep(3,8),1,1)),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
sf.str.out <- gsub("toprule",
"toprule 
  & \\\\multicolumn{4}{c}{\\\\textsc{thin}} & \\\\multicolumn{4}{c}{\\\\textsc{dense}} &&\\\\\\\\ 
\\\\cmidrule(lr{.5em}){2-5}\\\\cmidrule(lr{.5em}){6-9}
  & \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &
    \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &&\\\\\\\\
\\\\cmidrule(lr{.5em}){2-3}\\\\cmidrule(lr{.5em}){4-5}
\\\\cmidrule(lr{.5em}){6-7}\\\\cmidrule(lr{.5em}){8-9}", sf.str.out)
sf.str.out <- gsub("table","table*",sf.str.out)
clip <- pipe("pbcopy", "w")
write(sf.str.out, file=clip)                               
close(clip)
#cat(str.out)
```

And now the pitch data.

```{r}
# subsetting
dat.simple.pitch <- dat[dat$sim.type=="simple" & dat$data.type=="pitch",]
# aggregating
dat.simple.pitch.aggr <- 
  dat.simple.pitch[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                         par.type.1=mean(par.p[error.type=="type_1"] < 0.05),
                         smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                         par.power=mean(par.p[error.type=="type_2"] < 0.05),
                         time.med=median(time),
                         memory.med=median(memory)),
                      by=.(data.density,ranef,AR)]
kable(dat.simple.pitch.aggr)

# creating output table for formant data
# record timing / memory for dense data, set aside
sp.time.mem <- dat.simple.pitch.aggr[data.density=="dense",c("ranef","AR","time.med","memory.med")]
# go to wide format
sp.out <- dcast.data.table(dat.simple.pitch.aggr, ranef + AR ~ data.density, value.var=c("par.type.1","smooth.type.1","par.power","smooth.power"))
# reorder rows
sp.out <- sp.out[c(2:7,1),]
# merge back with timing / memory stuff
sp.out <- sp.out[sp.time.mem, on=.(ranef,AR)]
# new column that codes ranef and AR together
sp.out[,model.component:=ifelse(AR=="AR_est","AR1",ranef)]
sp.out$model.component <- recode(sp.out$model.component,
                                 noranef="no components",
                                 rintcpt="rand intcpt",
                                 rslope="rand intcpt + slope",
                                 rsmooth_tp_5="rand smooth, 5 bs",
                                 rsmooth_tp_8="rand smooth, 8 bs",
                                 rsmooth_tp_12="rand smooth, 12 bs")
sp.out <- sp.out[,-(1:2)]
setcolorder(sp.out, c("model.component",
                      "par.type.1_thin","par.power_thin",
                      "smooth.type.1_thin","smooth.power_thin",
                      "par.type.1_dense","par.power_dense",
                      "smooth.type.1_dense","smooth.power_dense",
                      "memory.med","time.med"))
colnames(sp.out) <- c("model component",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "memory","time")
sp.out <- as.data.frame(sp.out)
for (i in seq(2,8,2)) {
  sp.out[,i] <- as.character(round(sp.out[,i],3))
  sp.out[,i+1] <- as.character(round(sp.out[,i+1],3))
  for (j in 1:nrow(sp.out)) {
    if (as.numeric(sp.out[j,i]) > 0.15) {
      sp.out[j,i] <- paste0("\\color{lightgray} ",sp.out[j,i])
      sp.out[j,i+1] <- paste0("\\color{lightgray} ",sp.out[j,i+1])
    }
  }
}
sp.str.out <- print(xtable(sp.out, digits=c(0,0,rep(3,8),1,1)),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
sp.str.out <- gsub("toprule",
"toprule 
  & \\\\multicolumn{4}{c}{\\\\textsc{thin}} & \\\\multicolumn{4}{c}{\\\\textsc{dense}} &&\\\\\\\\ 
\\\\cmidrule(lr{.5em}){2-5}\\\\cmidrule(lr{.5em}){6-9}
  & \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &
    \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &&\\\\\\\\
\\\\cmidrule(lr{.5em}){2-3}\\\\cmidrule(lr{.5em}){4-5}
\\\\cmidrule(lr{.5em}){6-7}\\\\cmidrule(lr{.5em}){8-9}", sp.str.out)
sp.str.out <- gsub("table","table*",sp.str.out)
clip <- pipe("pbcopy", "w")
write(sp.str.out, file=clip)                               
close(clip)
```

## Random effects for capturing across-group patterns

`sim.type == "complex_across"`.

```{r}
dat.ca.formants <- dat[dat$sim.type=="complex_across" & dat$data.type=="formants",]

# aggregating data
dat.ca.formants.aggr <- 
  dat.ca.formants[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                         par.type.1=mean(par.p[error.type=="type_1"] < 0.05),
                         smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                         par.power=mean(par.p[error.type=="type_2"] < 0.05),
                         time.med=median(time),
                         memory.med=median(memory)),
                      by=.(data.density,ranef)]
kable(dat.ca.formants.aggr)

# creating output table for formant data
# record timing / memory for dense data, set aside
caf.time.mem <- dat.ca.formants.aggr[data.density=="dense",c("ranef","time.med","memory.med")]
# go to wide format
caf.out <- dcast.data.table(dat.ca.formants.aggr, ranef ~ data.density, value.var=c("par.type.1","smooth.type.1","par.power","smooth.power"))
# reorder rows
caf.out <- caf.out[c(1:3,5,6,4),]
# merge back with timing / memory stuff
caf.out <- caf.out[caf.time.mem, on=.(ranef)]
# new column that codes ranef and AR together
caf.out$model.component <- recode(unlist(lapply(strsplit(caf.out$ranef, split="+", fixed=T), function (x) {x[[2]]})),
                                 noranef="no random effect",
                                 rintcpt="rand intcpt",
                                 rslope="rand intcpt + slope",
                                 rsmooth_tp_3="rand smooth, 3 bs",
                                 rsmooth_tp_5="rand smooth, 5 bs",
                                 rsmooth_tp_10="rand smooth, 10 bs")
caf.out <- caf.out[,-1]
setcolorder(caf.out, c("model.component",
                      "par.type.1_thin","par.power_thin",
                      "smooth.type.1_thin","smooth.power_thin",
                      "par.type.1_dense","par.power_dense",
                      "smooth.type.1_dense","smooth.power_dense",
                      "memory.med","time.med"))
colnames(caf.out) <- c("random effects",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "memory","time")
caf.out <- as.data.frame(caf.out)
for (i in seq(2,8,2)) {
  caf.out[,i] <- as.character(round(caf.out[,i],3))
  caf.out[,i+1] <- as.character(round(caf.out[,i+1],3))
  for (j in 1:nrow(caf.out)) {
    if (as.numeric(caf.out[j,i]) > 0.15) {
      caf.out[j,i] <- paste0("\\color{lightgray} ",caf.out[j,i])
      caf.out[j,i+1] <- paste0("\\color{lightgray} ",caf.out[j,i+1])
    }
  }
}
caf.str.out <- print(xtable(caf.out, digits=c(0,0,rep(3,8),1,1)),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
caf.str.out <- gsub("toprule",
"toprule 
  & \\\\multicolumn{4}{c}{\\\\textsc{thin}} & \\\\multicolumn{4}{c}{\\\\textsc{dense}} &&\\\\\\\\ 
\\\\cmidrule(lr{.5em}){2-5}\\\\cmidrule(lr{.5em}){6-9}
  & \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &
    \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &&\\\\\\\\
\\\\cmidrule(lr{.5em}){2-3}\\\\cmidrule(lr{.5em}){4-5}
\\\\cmidrule(lr{.5em}){6-7}\\\\cmidrule(lr{.5em}){8-9}", caf.str.out)
caf.str.out <- gsub("table","table*",caf.str.out)
clip <- pipe("pbcopy", "w")
write(caf.str.out, file=clip)                               
close(clip)

```


```{r}
dat.ca.pitch <- dat[dat$sim.type=="complex_across" & dat$data.type=="pitch",]
dat.ca.pitch.aggr <- 
  dat.ca.pitch[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                         par.type.1=mean(par.p[error.type=="type_1"] < 0.05),
                         smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                         par.power=mean(par.p[error.type=="type_2"] < 0.05),
                         time.med=median(time),
                         memory.med=median(memory)),
                      by=.(data.density,ranef)]
kable(dat.ca.pitch.aggr)

# creating output table for formant data
# record timing / memory for dense data, set aside
cap.time.mem <- dat.ca.pitch.aggr[data.density=="dense",c("ranef","time.med","memory.med")]
# go to wide format
cap.out <- dcast.data.table(dat.ca.pitch.aggr, ranef ~ data.density, value.var=c("par.type.1","smooth.type.1","par.power","smooth.power"))
# reorder rows
cap.out <- cap.out[c(1:3,5,6,4),]
# merge back with timing / memory stuff
cap.out <- cap.out[cap.time.mem, on=.(ranef)]
# new column that codes ranef and AR together
cap.out$model.component <- recode(unlist(lapply(strsplit(cap.out$ranef, split="+", fixed=T), function (x) {x[[2]]})),
                                 noranef="no random effect",
                                 rintcpt="rand intcpt",
                                 rslope="rand intcpt + slope",
                                 rsmooth_tp_5="rand smooth, 5 bs",
                                 rsmooth_tp_8="rand smooth, 8 bs",
                                 rsmooth_tp_12="rand smooth, 12 bs")
cap.out <- cap.out[,-1]
setcolorder(cap.out, c("model.component",
                      "par.type.1_thin","par.power_thin",
                      "smooth.type.1_thin","smooth.power_thin",
                      "par.type.1_dense","par.power_dense",
                      "smooth.type.1_dense","smooth.power_dense",
                      "memory.med","time.med"))
colnames(cap.out) <- c("random effects",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "memory","time")
cap.out <- as.data.frame(cap.out)
for (i in seq(2,8,2)) {
  cap.out[,i] <- as.character(round(cap.out[,i],3))
  cap.out[,i+1] <- as.character(round(cap.out[,i+1],3))
  for (j in 1:nrow(cap.out)) {
    if (as.numeric(cap.out[j,i]) > 0.15) {
      cap.out[j,i] <- paste0("\\color{lightgray} ",cap.out[j,i])
      cap.out[j,i+1] <- paste0("\\color{lightgray} ",cap.out[j,i+1])
    }
  }
}
cap.str.out <- print(xtable(cap.out, digits=c(0,0,rep(3,8),1,1)),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
cap.str.out <- gsub("toprule",
"toprule 
  & \\\\multicolumn{4}{c}{\\\\textsc{thin}} & \\\\multicolumn{4}{c}{\\\\textsc{dense}} &&\\\\\\\\ 
\\\\cmidrule(lr{.5em}){2-5}\\\\cmidrule(lr{.5em}){6-9}
  & \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &
    \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &&\\\\\\\\
\\\\cmidrule(lr{.5em}){2-3}\\\\cmidrule(lr{.5em}){4-5}
\\\\cmidrule(lr{.5em}){6-7}\\\\cmidrule(lr{.5em}){8-9}", cap.str.out)
cap.str.out <- gsub("table","table*",cap.str.out)
clip <- pipe("pbcopy", "w")
write(cap.str.out, file=clip)                               
close(clip)
```
## Random effects for capturing within-group patterns

`sim.type == "complex_within"`.

```{r}
dat.cw.formants <- dat[dat$sim.type=="complex_within" & dat$data.type=="formants",]
# aggregating data
dat.cw.formants.aggr <- 
  dat.cw.formants[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                         par.type.1=mean(par.p[error.type=="type_1"] < 0.05),
                         smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                         par.power=mean(par.p[error.type=="type_2"] < 0.05),
                         time.med=median(time),
                         memory.med=median(memory)),
                      by=.(data.density,ranef)]
kable(dat.cw.formants.aggr)

# creating output table for formant data
# record timing / memory for dense data, set aside
cwf.time.mem <- dat.cw.formants.aggr[data.density=="dense",c("ranef","time.med","memory.med")]
# go to wide format
cwf.out <- dcast.data.table(dat.cw.formants.aggr, ranef ~ data.density, value.var=c("par.type.1","smooth.type.1","par.power","smooth.power"))
# reorder rows
cwf.out <- cwf.out[c(2,3,1,8,9,7,5,6,4),]
# merge back with timing / memory stuff
cwf.out <- cwf.out[cwf.time.mem, on=.(ranef)]
# new column that codes ranef and AR together
cwf.out$model.component <- recode(unlist(lapply(strsplit(cwf.out$ranef, split="+", fixed=T), function (x) {x[[2]]})),
                                 `rsmooth_tp_3`="r smooth 3 $|$ wrd",
                                 `rsmooth_tp_5`="r smooth 5 $|$ wrd",
                                 `rsmooth_tp_10`="r smooth 10 $|$ wrd",
                                 `rsmoothslope_tp_3`="r slope \\& smooth 3 $|$ wrd",
                                 `rsmoothslope_tp_5`="r slope \\& smooth 5 $|$ wrd",
                                 `rsmoothslope_tp_10`="r slope \\& smooth 10 $|$ wrd",
                                 `rsmoothcrossed_tp_3`="r smooth 3 $|$ wrd$\\times$grp",
                                 `rsmoothcrossed_tp_5`="r smooth 5 $|$ wrd$\\times$grp",
                                 `rsmoothcrossed_tp_10`="r smooth 10 $|$ wrd$\\times$grp"
                                 )
cwf.out <- cwf.out[,-1]
setcolorder(cwf.out, c("model.component",
                      "par.type.1_thin","par.power_thin",
                      "smooth.type.1_thin","smooth.power_thin",
                      "par.type.1_dense","par.power_dense",
                      "smooth.type.1_dense","smooth.power_dense",
                      "memory.med","time.med"))
colnames(cwf.out) <- c("random effects",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "type I","power",
                      "memory","time")
cwf.out <- as.data.frame(cwf.out)
for (i in seq(2,8,2)) {
  cwf.out[,i] <- as.character(round(cwf.out[,i],3))
  cwf.out[,i+1] <- as.character(round(cwf.out[,i+1],3))
  for (j in 1:nrow(cwf.out)) {
    if (as.numeric(cwf.out[j,i]) > 0.15) {
      cwf.out[j,i] <- paste0("\\color{lightgray} ",cwf.out[j,i])
      cwf.out[j,i+1] <- paste0("\\color{lightgray} ",cwf.out[j,i+1])
    }
  }
}
cwf.str.out <- print(xtable(cwf.out, digits=c(0,0,rep(3,8),1,1)),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
cwf.str.out <- gsub("toprule",
"toprule 
  & \\\\multicolumn{4}{c}{\\\\textsc{thin}} & \\\\multicolumn{4}{c}{\\\\textsc{dense}} &&\\\\\\\\ 
\\\\cmidrule(lr{.5em}){2-5}\\\\cmidrule(lr{.5em}){6-9}
  & \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &
    \\\\multicolumn{2}{c}{\\\\textsc{parametric}} & \\\\multicolumn{2}{c}{\\\\textsc{smooth}} &&\\\\\\\\
\\\\cmidrule(lr{.5em}){2-3}\\\\cmidrule(lr{.5em}){4-5}
\\\\cmidrule(lr{.5em}){6-7}\\\\cmidrule(lr{.5em}){8-9}", cwf.str.out)
cwf.str.out <- gsub("table","table*",cwf.str.out)
clip <- pipe("pbcopy", "w")
write(cwf.str.out, file=clip)                               
close(clip)

```

## Different methods of significance testing

`sim.type == "complex_sigtest"`.

```{r}
make.table <- function (tbl, selectors) {
  tbl <- as.data.frame(tbl, stringsAsFactors=F)
  out <- data.frame(type.of.test="", type.1=0, power=0, memory=0, time=0, stringsAsFactors=F)[0,]
  summarycol <- paste(tbl$diff.smooth.type,tbl$method,sep=",")
  for (s in 1:length(selectors)) {
    sel <- selectors[[s]]
    out[s,] <- list(sel$name,
                    tbl[summarycol==sel$which.row,paste(sel$which.p,"type.1",sep=".")],
                    tbl[summarycol==sel$which.row,paste(sel$which.p,"power",sep=".")],
                    tbl[summarycol==sel$which.row,"memory.med"],
                    tbl[summarycol==sel$which.row,"time.med"])
    print(out[s,])
  }
  return(out)
}

make.vis.table <- function (tbl, selectors) {
  tbl <- as.data.frame(tbl, stringsAsFactors=F)
  out <- data.frame(type.of.test="", type.1=0, power=0, type.1=0, power=0, stringsAsFactors=F)[0,]
  summarycol <- paste(tbl$diff.smooth.type,tbl$method,sep=",")
  for (s in 1:length(selectors)) {
    sel <- selectors[[s]]
    out[s,] <- list(sel$name,
                    tbl[summarycol==sel$which.row,paste(sel$which.p,"type.1",sep=".")],
                    tbl[summarycol==sel$which.row,paste(sel$which.p,"power",sep=".")],
                    tbl[summarycol==sel$which.row,"memory.med"],
                    tbl[summarycol==sel$which.row,"time.med"])
    print(out[s,])
  }
  return(out)
}


dat.cs.formants <- dat[dat$sim.type=="complex_sigtest" & dat$data.type=="formants",]

# aggregating data
dat.cs.formants.aggr <- 
  dat.cs.formants[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                     both.type.1=mean(par.p[error.type=="type_1"] < 0.05 | smooth.p[error.type=="type_1"] < 0.05),
                     bonferroni.type.1=mean(par.p[error.type=="type_1"] < 0.025 | smooth.p[error.type=="type_1"] < 0.025),
                     mod.comp.type.1=mean(mod.comp.p[error.type=="type_1"] < 0.05),
                     smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                     both.power=mean(par.p[error.type=="type_2"] < 0.05 | smooth.p[error.type=="type_2"] < 0.05),
                     bonferroni.power=mean(par.p[error.type=="type_2"] < 0.025 | smooth.p[error.type=="type_2"] < 0.025),
                     mod.comp.power=mean(mod.comp.p[error.type=="type_2"] < 0.05),
                     time.med=median(time),
                     memory.med=median(memory)),
                  by=.(diff.smooth.type,method)]
kable(dat.cs.formants.aggr)

csf.out <- make.table(dat.cs.formants.aggr,
                     list(
                       list(name="par/smooth",
                            which.row="diff_tp_10,discrete",
                            which.p="both"),
                       list(name="par/smooth bonf.",
                            which.row="diff_tp_10,discrete",
                            which.p="bonferroni"),
                       list(name="LRT, ML",
                            which.row="diff_tp_10,ML",
                            which.p="mod.comp"),
                       list(name="LRT, fREML",
                            which.row="diff_tp_10,discrete",
                            which.p="mod.comp"),
                       list(name="binary smooth",
                            which.row="bin_tp_10,discrete",
                            which.p="smooth"),
                       list(name="LRT, fREML trick",
                            which.row="diff_tp_10,discrete+select",
                            which.p="mod.comp")
                     ))

colnames(csf.out) <- c("significance test",
                      "type I","power",
                      "memory","time")
csf.out <- as.data.frame(csf.out)
csf.out[,2] <- as.character(round(csf.out[,2],3))
csf.out[,3] <- as.character(round(csf.out[,3],3))
for (j in 1:nrow(csf.out)) {
  if (as.numeric(csf.out[j,2]) > 0.15) {
    csf.out[j,2] <- paste0("\\color{lightgray} ",csf.out[j,2])
    csf.out[j,3] <- paste0("\\color{lightgray} ",csf.out[j,3])
  }
}

csf.str.out <- print(xtable(csf.out, digits=c(0,0,0,0,1,1)),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
csf.str.out <- gsub("lllrr", "ll@{\\\\hspace{0.7em}}l@{\\\\hspace{0.7em}}r@{\\\\hspace{0.7em}}r", csf.str.out)
clip <- pipe("pbcopy", "w")
write(csf.str.out, file=clip)                               
close(clip)

# no table for visual data; instead, an ecdf plot is used

dat.cs.formants.vis.type.1 <- dat.cs.formants[dat.cs.formants$method=="ML" & dat.cs.formants$error.type=="type_1",]
dat.cs.formants.vis.power <- dat.cs.formants[dat.cs.formants$method=="ML" & dat.cs.formants$error.type=="type_2",]
reverse.ecdf <- function (x) {
  Vectorize(function (y) {mean(x >= y)}, "y")
}
csf.overlap.type.1 <- reverse.ecdf(dat.cs.formants.vis.type.1$visual.no.overlap)(seq(0.01,1,0.01))
csf.overlap.power <- reverse.ecdf(dat.cs.formants.vis.power$visual.no.overlap)(seq(0.01,1,0.01))
csf.diff.type.1 <- reverse.ecdf(dat.cs.formants.vis.type.1$visual.excludes.0)(seq(0.01,1,0.01))
csf.diff.power <- reverse.ecdf(dat.cs.formants.vis.power$visual.excludes.0)(seq(0.01,1,0.01))

pdf("../jphon/images/f2_visual.pdf",width=6, height=4.5)
plot(c(), c(), xlim=c(0,100), ylim=c(0,1), xlab="", ylab="", cex.axis=1.25,
     xaxt="n")
axis(1, at=c(1,20,40,60,80,100), cex.axis=1.25)
title(xlab="Cut-off %", cex.lab=1.25, font.lab=2)
title(ylab="Type I error rate / Power", cex.lab=1.25, font.lab=2)
title(main="Visual methods: F2 contours", cex.lab=1.25, font.lab=2, cex.main=1.75)
lines(seq(1,100,1), csf.overlap.type.1, lwd=3, col="firebrick3", lty=2)
lines(seq(1,100,1), csf.overlap.power, lwd=3, col="deepskyblue4", lty=2)
lines(seq(1,100,1), csf.diff.type.1, lwd=3, col="firebrick3", lty=1)
lines(seq(1,100,1), csf.diff.power, lwd=3, col="deepskyblue4", lty=1)
abline(h=0.05, col="gray", lty=3, lw=2)
legend("topright", c("overlap, type I","overlap, power", 
                     "difference, type I", "difference, power"),
       col=c("firebrick3","deepskyblue4","firebrick3","deepskyblue4"),
       lty=c(2,2,1,1), lwd=rep(2,4))
dev.off()


```

And now the pitch contours.

```{r}
dat.cs.pitch <- dat[dat$sim.type=="complex_sigtest" & dat$data.type=="pitch",]
# aggregating data
dat.cs.pitch.aggr <- 
  dat.cs.pitch[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                     both.type.1=mean(par.p[error.type=="type_1"] < 0.05 | smooth.p[error.type=="type_1"] < 0.05),
                     bonferroni.type.1=mean(par.p[error.type=="type_1"] < 0.025 | smooth.p[error.type=="type_1"] < 0.025),
                     mod.comp.type.1=mean(mod.comp.p[error.type=="type_1"] < 0.05),
                     smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                     both.power=mean(par.p[error.type=="type_2"] < 0.05 | smooth.p[error.type=="type_2"] < 0.05),
                     bonferroni.power=mean(par.p[error.type=="type_2"] < 0.025 | smooth.p[error.type=="type_2"] < 0.025),
                     mod.comp.power=mean(mod.comp.p[error.type=="type_2"] < 0.05),
                     time.med=median(time),
                     memory.med=median(memory)),
                  by=.(diff.smooth.type,method)]
kable(dat.cs.formants.aggr)

csp.out <- make.table(dat.cs.pitch.aggr,
                     list(
                       list(name="par/smooth",
                            which.row="diff_tp_15,discrete",
                            which.p="both"),
                       list(name="par/smooth bonf.",
                            which.row="diff_tp_15,discrete",
                            which.p="bonferroni"),
                       list(name="LRT, ML",
                            which.row="diff_tp_15,ML",
                            which.p="mod.comp"),
                       list(name="LRT, fREML",
                            which.row="diff_tp_15,discrete",
                            which.p="mod.comp"),
                       list(name="binary smooth",
                            which.row="bin_tp_15,discrete",
                            which.p="smooth"),
                       list(name="LRT, fREML trick",
                            which.row="diff_tp_15,discrete+select",
                            which.p="mod.comp")
                     ))

colnames(csp.out) <- c("significance test",
                      "type I","power",
                      "memory","time")
csp.out <- as.data.frame(csp.out)
csp.out[,2] <- as.character(round(csp.out[,2],3))
csp.out[,3] <- as.character(round(csp.out[,3],3))
for (j in 1:nrow(csp.out)) {
  if (as.numeric(csp.out[j,2]) > 0.15) {
    csp.out[j,2] <- paste0("\\color{lightgray} ",csp.out[j,2])
    csp.out[j,3] <- paste0("\\color{lightgray} ",csp.out[j,3])
  }
}

csp.str.out <- print(xtable(csp.out, digits=c(0,0,0,0,1,1)),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
csp.str.out <- gsub("lllrr", "ll@{\\\\hspace{0.7em}}l@{\\\\hspace{0.7em}}r@{\\\\hspace{0.7em}}r", csp.str.out)
clip <- pipe("pbcopy", "w")
write(csp.str.out, file=clip)                               
close(clip)

dat.cs.pitch.vis.type.1 <- dat.cs.pitch[dat.cs.pitch$method=="ML" & dat.cs.pitch$error.type=="type_1",]
dat.cs.pitch.vis.power <- dat.cs.pitch[dat.cs.pitch$method=="ML" & dat.cs.pitch$error.type=="type_2",]
reverse.ecdf <- function (x) {
  Vectorize(function (y) {mean(x >= y)}, "y")
}
csp.overlap.type.1 <- reverse.ecdf(dat.cs.pitch.vis.type.1$visual.no.overlap)(seq(0.01,1,0.01))
csp.overlap.power <- reverse.ecdf(dat.cs.pitch.vis.power$visual.no.overlap)(seq(0.01,1,0.01))
csp.diff.type.1 <- reverse.ecdf(dat.cs.pitch.vis.type.1$visual.excludes.0)(seq(0.01,1,0.01))
csp.diff.power <- reverse.ecdf(dat.cs.pitch.vis.power$visual.excludes.0)(seq(0.01,1,0.01))

pdf("../jphon/images/pitch_visual.pdf",width=6, height=4.5)
plot(c(), c(), xlim=c(0,100), ylim=c(0,1), xlab="", ylab="", cex.axis=1.25,
     xaxt="n")
axis(1, at=c(1,20,40,60,80,100), cex.axis=1.25)
title(xlab="Cut-off %", cex.lab=1.25, font.lab=2)
title(ylab="Type I error rate / Power", cex.lab=1.25, font.lab=2)
title(main="Visual methods: Pitch contours", cex.lab=1.25, font.lab=2, cex.main=1.75)
lines(seq(1,100,1), csp.overlap.type.1, lwd=3, col="firebrick3", lty=2)
lines(seq(1,100,1), csp.overlap.power, lwd=3, col="deepskyblue4", lty=2)
lines(seq(1,100,1), csp.diff.type.1, lwd=3, col="firebrick3", lty=1)
lines(seq(1,100,1), csp.diff.power, lwd=3, col="deepskyblue4", lty=1)
abline(h=0.05, col="gray", lty=3, lw=2)
legend("topright", c("overlap, type I","overlap, power", 
                     "difference, type I", "difference, power"),
       col=c("firebrick3","deepskyblue4","firebrick3","deepskyblue4"),
       lty=c(2,2,1,1), lwd=rep(2,4))
dev.off()
```
